---
title: "R Notebook"
output: html_notebook
---

# R Basic

## Introduction

### R as a calculator

```{r}
1 + 100
```

```{r}
2^2
```

```{r}
7/2
```

```{r}
4-1
```

```{r}
2*2
```

```{r}
3+5*2
```

```{r}
(3+5)*2
```

```{r}
2e2
```

```{r}
2/10000
```

```{r}
sin(1)
```

```{r}
log(1)
```

```{r}
exp(0)
```

```{r}
1 == 1
```

```{r}
1 >= 1
```

```{r}
1 < 1
```

```{r}
1 != 2
```

```{r}
x <- 10
x
```

```{r}
x + x
```

```{r}
x - 5
```

```{r}
x^2
```

```{r}
y <- x + x
y
```

### Vectorization

```{r}
1:5
```

```{r}
2^(1:5)
```

```{r}
v <- 1:5
2^v
```

```{r}
log10(v)
```

### Environment

```{r}
ls()
```
```{r}
rm(v)
ls()
```

## Package Management

```{r}
installed.packages()
```

```{r}
#install.packages("vegan", dependencies = TRUE) #Already installed
```

```{r}
library(vegan)
```


```{r}
#remove.packages()
```

## Project Management

1. Set up management structure (i.e., "data", "src", "results", "doc")
2. Discuss avoiding redundancy of files
3. Make sure to use version control (i.e., Git)

## Getting Help

```{r}
?plot
```

```{r}
help(plot)
```

```{r}
?"<-"
```

```{r}
vignette("FAQ-vegan")
```

```{r}
citation("vegan")
```

## Data and their formats

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                   weight = c(2.1,5,3.2),
                   likes_string = c(1,0,1))
```

```{r}
cats
```

```{r}
write.csv(cats, file = "~/GitHub/biol48006220_Fall2022/practcomp_2022/data/cats.csv")
```

```{r}
cats$coat
```

```{r}
cats$weight
```

```{r}
cats$weight*10
```

```{r}
log(cats$weight)
```

```{r}
logweight <- log(cats$weight)
cbind(cats, logweight)
```

```{r}
cats <- cbind(cats, logweight)
```

```{r}
paste("My cat is ", cats$coat, ", and it weighs ", cats$weight, " kg.", sep = "")
```

### Data Types

```{r}
typeof(cats$coat)
```

```{r}
typeof(cats$weight)
```

```{r}
typeof(cats$likes_string)
```

```{r}
typeof(3.14159)
```

```{r}
typeof(1i)
```

```{r}
typeof(FALSE)
```

```{r}
typeof(T)
```

```{r}
typeof(cats)
```

```{r}
class(cats)
```

```{r}
file.show("./data/cats.csv")
```

```{r}
cats$likes_string <- as.logical(cats$likes_string)
typeof(cats$likes_string)
```

```{r}
cats$likes_string
```

```{r}
c("a", "b")
```

```{r}
ab <- c("a", "b")
ab
```

```{r}
c(ab, "c")
```

```{r}
c(ab, ab)
```

```{r}
1:10
```

```{r}
seq(10)
```

```{r}
z <- seq(10)
head(z, n=3)
```

```{r}
length(z)
```

```{r}
class(z)
```

```{r}
typeof(z)
```

```{r}
seq(0,100, by=5)
```

### Data Frames
```{r}
cats
```

```{r}
str(cats$coat)
```

```{r}
coats <- c("tabby", "tortoiseshell", "tortoiseshell", "black", "tabby")
coats
```

```{r}
str(coats)
```

```{r}
factor(coats)
```

```{r}
categories <- factor(coats)
class(coats)
class(categories)
```

```{r}
str(categories)
```

### Lists

```{r}
list_example <- list(title="Numbers", numbers = 1:10, data=T)
```

```{r}
list_example
```

```{r}
another_list <- list(1, "a", TRUE, 1+1i)
another_list
```

```{r}
typeof(list_example)
class(list_example)
```

```{r}
typeof(cats)
class(cats)
```

```{r}
data.frame(list_example)
```

```{r}
cats
cats[,3]
cats[3,]
```

```{r}
cats[2:3, c(1,3)]
```

### Matrices


```{r}
matrix_example<- matrix(0,ncol=5, nrow=3)
```
```{r}
matrix_example

```


```{r}
class(matrix_example)
```
```{r}
typeof(matrix_example)
```
```{r}
str(matrix_example)
```
```{r}
dim(matrix_example)
ncol(matrix_example)
nrow(matrix_example)
```

the dim() function will tell you info on the shape

```{r}
class(data.frame(matrix_example))
```
This will make it in to a df
```{r}
df_example<- data.frame(matrix_example)
df_example
```
now it's saved as a df

### Subsetting

```{r}
p <-c(2.3, 6.9,4.0,23,1)
p
```
```{r}
names(p)<- c('a','b','c','d','e')
p
```

```{r}
p[1]
```
```{r}
p[2:4]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[6]
```
out of range items will not crash your script
```{r}
p[-3]
```
the above code gives you everything besides 3

```{r}
p[-(2:4)]
```

above code is how you drop a range

```{r}
p[c('a','c')]
```
```{r}
p[c(T,F,T,T,F)]
```
 T = give data point F= leave out data point
 
```{r}
p[names(p) !='c']
```
 
This leaves out p

### Factors

```{r}
f <- factor(c('a','b','c','d','e'))
f
```
```{r}
f[f=='a']
```

Above code gives you f where it is equal to a. This can be useful in for loops

```{r}
f[1:3]
```

```{r}
f[f %in% c('b','c')]
```
```{r}
f[-3]
```

```{r}
f2<- c('a','a','b','c','c')
f2
```
```{r}
f2[f2=='a']
```

```{r}
f2[f2 %in% c('a','c')]
```
%in% means "search in here for this"


### matrices resumed
```{r}
set.seed(65)
m <- matrix(rnorm(6*4),ncol=4, nrow=6)
m

```
#set seed ensures consistency when "randomly" generating seeds. People can reproduce your results

```{r}
m[3:4, c(3,1)]
```
row 3-4, columns 3 to 1


###lists revisited

```{r}
xlist <-list(a="BIOL48006220", b=seq(1,10, by = 0.5), data = "Grade")
xlist
```




```{r}
xlist[1:2]
```

```{r}
xlist['a']
```

```{r}
xlist['b']
```
##gives as a sublist
```{r}
xlist[['b']]
```
the double bracket is used to call in
```{r}
xlist$b
```

### Data Frames revisited
```{r}
getwd()
```
```{r}
gp <- read.csv('./data/forest_area_sq_km.csv')
head(gp, n=10L)
```

```{r}
head(gp[3], n=10L)
```

```{r}
nrow(gp)
```
```{r}
head(gp[["ï..country"]],10L)
```
```{r}
gp$X2001
```

```{r}
gp[1:3, 2:5]
```
row, columns

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100 ), ]
```

### Conditionals and Flow
```{r}
n <- 10

if (n<10) {
  print("n is less than 10")
} else if (n >10 ){
  print("n is greater than 10")
} else {
  print("n is equal to 10")
}
```

```{r}
for (n in seq(1,20)){
  if (n<10) {
  print("n is less than 10")
} else if (n >10 ){
  print("n is greater than 10")
} else {
  print("n is equal to 10")
}
}

```

```{r}
g<- 0

while(g <= 10){
  print(paste(g, "is less than or equal to 10"))
  g <- g+1
}
```
## Plotting and Figures
```{r}
#install.packages("ggplot2")
```
```{r}
library(ggplot2)
```


```{r}
forestdata<-read.csv('./data/forest_area_sq_km.csv')
```

```{r}
ggplot(data=forestdata, mapping=aes(x=X1989, y=X2019))+geom_point()
```


```{r}
gp <-read.csv('./data/gapminder_all.csv')

head(gp)
```
```{r}
ggplot(data=gp, mapping=aes(x=gdpPercap_1952, y=pop_1952 )) + geom_point()
```
```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv")

head(gapminder)
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country, color=continent)) + geom_line()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping=aes(color=continent))+geom_point()
```
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp))+ geom_point() 
```
# this data needs to be log transformed 
```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp))+ geom_point(alpha=0.25) + scale_x_log10()
```

Now you can run linear regression model
alpha= transparency of data point

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp))+ geom_point(alpha=0.25, color="purple") + scale_x_log10()+ geom_smooth(method=lm, color="gold", size= 1.25)
```
```{r}
africas<- gapminder[gapminder$continent=="Africa",]
head(africas)
```
```{r}
ggplot(data=africas, mapping= aes(x=year, y=lifeExp))+geom_line()+ facet_wrap( ~country) + theme(axis.text.x= element_text(angle=90)) +
  labs(
    x= "year",
    y="life expectancy",
    title= "Life Expectancy Over Time in African Countries"
  )
                                                                                                                        
```
```{r}
AfricanLifeExp <- ggplot(data=africas, mapping= aes(x=year, y=lifeExp))+geom_line()+ facet_wrap( ~country) + theme(axis.text.x= element_text(angle=90)) +
  labs(
    x= "year",
    y="life expectancy",
    title= "Life Expectancy Over Time in African Countries"
  )
                                                                        
```

now your graph is saved as an object
```{r}
ggsave(filename= "data/AfricanLifeExp.png", plot =AfricanLifeExp, width = 24, height = 40, dpi =300, units="cm")
```

```{r}
pdf(file= "results/AfricanLifeExp.pdf", width = 24, height = 40)
plot(AfricanLifeExp)
dev.off()
```
 must turn developer mode off, this is how you get figures out of R
 
```{r}
write.table(gapminder, file="data/gapminder_web.csv", sep=",")
```
 
```{r}
write.csv(africas, file="data/gapminder_web_africas.csv")
```


```{r}
install.packages(c("gridges","viridis", "hrbrthemes"), dependencies = T)
```
r graph gallery

## Stats

### Additional ways of importing the data

```{r}
rand<- c(12,54,98,65,38)
```


```{r}
rand
```
```{r}
```


```{r}
sum(rand)
```
```{r}
length(rand)
avg(rand)
mean(rand)
sort(rand)
min(rand)
max(rand)
cumsum(rand)
diff(rand)
rand[2]
rand*2
```


```{r}
pedes <- scan()
```

^^can be used to copy and paste data into a df

### Analyses

#### Univariate Statistics
1. Categorical data
  a. Barplots
```{r}
beer <- c(3,4,1,1,3,4,3,1,3,2,1,2,1,2,1,1,1,4,3,1)
```

  
```{r}
barplot(table(beer), xlab="beer", ylab="Frequency") #counts
```
  
```{r}
barplot(table(beer)/length(beer), xlab)

#unfinished
```
  
```{r}
pie(table(beer), main="beer preference")
```
  
2.Numerical data
  a. stem and leaf plots
```{r}
stem(pedes)
```

```{r}
stripchart(pedes,method="stack")
```

3. Measures of Center
  a. mean
```{r}
mean(pedes)
```
b. Median
```{r}
median(pedes)
median(gapminder$country)
```
```{r}
with(gapminder, median(lifeExp))
```

c.mode

```{r}
which(table(pedes)==max(table(pedes)))
```

d. variation
  a.range
  
```{r}
range(pedes)
```
b. variance
```{r}
var(pedes)
```
IqR
```{r}
IQR(pedes)
```

d. z scores
```{r}
scale(pedes)
```

what are z score- # of SD away from mean

e. Summary
```{r}
Summary(pedes)
```

histogram
```{r}
hist(pedes)
```

5. Plotting
a. histograms
```{r}
hist(pedes, breaks="scott")
#scott creates a nice number of breaks

hist(pedes, breaks="scott", prob=TRUE)
#will plot probability

hist(pedes, breaks="scott", prob=TRUE)
lines(density(pedes))
```

box plots
```{r}
boxplot(pedes)

summary(boxplot(pedes))
#aka
summary(pedes)
```


#### Bivariate and Multivariate Statistics

1. plotting and regression

```{r}
spid.gen <- read.csv("data/spider_genitalia.csv", header = T )
```

```{r}
boxplot(spid.gen$left.bulb ~ spid.gen$right.bulb)

#better shown as scatterplot
```

```{r}
boxplot(spid.gen$left.bulb ~ spid.gen$habitat)

boxplot(spid.gen$left.bulb/spid.gen$carapace ~ spid.gen$habitat)
```

gen.reg function will give you correlation info

```{r}
plot(spid.gen$left.bulb ~ spid.gen$right.bulb)
#abline(gen.reg)
```
d. Correlation Coefficients & Spearman Rank Correlation
```{r}
cor.gen <- with(spid.gen, cor(left.bulb,right.bulb))
cor.gen
cor.gen^2
```

cor is correlation, variance, and covariance in  a matrix fashion

This is your R value, you then square the value for r2.

```{r}
spearman.cor.gen <- with(spid.gen, cor(left.bulb, right.bulb, method="spearman"))
spearman.cor.gen
```
e. Residuals

```{r}
residuals(gen.reg)
```
f. transformations

```{r}
plot(spid.gen$left.bulb^2~spid.gen$right.bulb^2)
plot(log10(spid.gen$left.bulb)~(spid.gen$right.bulb))
```
2. Comparing discrete treatment effects
  a. chi-squared
  
```{r}
obs_weighted <- c(4,15,6,15,18,2)
obs_fair <- c(10,10,10,10,10,10)
exp <- c(.16,.17,.16,.17,.17,.17)
chisq.test(obs_weighted, p=exp)
chisq.test(obs_fair, p=exp)
```
  b. T-tests
```{r}
#H0- true mean is equal to 0, but you can set true mean to other values with mu=
t.test(spid.gen$carapace.length)

mean(spid.gen$carapace.length, mu=29)
```
```{r}
habitat.t2<-t.test(spid.gen$carapace.length~spid.gen$habitat)
habitat.t2
```
```{r}
hab.tless <- t.test(spid.gen$left.bulb, alternative = "greater")
hab.tless
```
```{r}
hab.tgreater <- t.test(spid.gen$left.bulb, alternative = "greater")
hab.tgreater
```
c. Analysis of Variance, ANOVA

compares mean between three or more groups . Its like doing all possible t-test

```{r}
gen.lm <- lm(spid.gen$left.bulb~spid.gen$habitat)
gen.lm
summary(gen.lm)
```
```{r}
anova(gen.lm)
```
```{r}
gen.anova <- aov(spid.gen$left.bulb~spid.gen$habitat)
summary(gen.anova)
```

```{r}
gapminder.lifeexp.cont.anova<-aov(gapminder$lifeExp~gapminder$continent)
summary(gapminder.lifeexp.cont.anova)
```
  
```{r}
gapminder.tukey.two.way<- TukeyHSD(gapminder.lifeexp.cont.anova)
gapminder.tukey.two.way

```
analysis of co-variance (ancova)




```{r}
plot(spid.gen$left.bulb~spid.gen$habitat)
```
```{r}
gen.ancova <- lm(spid.gen$left.bulb~spid.gen$habitat*spid.gen$carapace.length)
summary(gen.ancova)
```
```{r}
gm.lifeexp_gdp.ancova<-lm(gapminder$lifeExp~gapminder$continent*gapminder$gdpPercap)
summary(gm.lifeexp_gdp.ancova)
```

```{r}
gm.lifeexp.cont.gdp.ancova<-aov(gapminder$lifeExp~gapminder$continent+gapminder$gdpPercap)
```

3. Principle component analysis
```{r}
sp.matrix <- with(spid.gen,cbind(left.bulb,right.bulb,carapace.length, leg4.length))
sp.matrix
```
```{r}
sp.pca<-princomp(sp.matrix, cor=TRUE)
summary(sp.pca)
loadings(sp.pca)
```
```{r}
biplot(sp.pca)
```


```{r}
```
```{r}
install.packages("ggfortify",dependencies = T)
library(ggfortify)
sp.pca.plot <- autoplot(sp.pca, 
                        data = spid.gen,
                        colour= 'habitat')
sp.pca.plot
```
```{r}
plot (1:25, rep(0.25,25), pch=1:25, col=1:25, ylim=c(0,6), cex=2, ylab="line types 1 to 6", xlab= "plotting character 1 to 25 and colors 1 to 8", main= "line types(lty), plotting characters(pch), |nand colors (col) for plot and xyplot", lab=c(25,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=3)

abline(h=1:6, lty=1:6, col=1:6, lwd=2)

#lwd is width
```
```{r}
pdf("./results/PlotCheatSheet.pdf")
plot (1:25, rep(0.25,25), pch=1:25, col=1:25, ylim=c(0,6), cex=2, ylab="line types 1 to 6", xlab= "plotting character 1 to 25 and colors 1 to 8", main= "line types(lty), plotting characters(pch), |nand colors (col) for plot and xyplot", lab=c(25,7,2))

points(1:8, rep(0.5,8), pch=20, col=1:8, cex=3)

abline(h=1:6, lty=1:6, col=1:6, lwd=2)

dev.off()
```
this wil save everything from the pdf line to dev.off() in a pdf. dev.off means developer mode off.


##Looping and conditonals

a. if then statements
```{r}
x <- -5
if(x > 0 ){
  print ("Positive Number")
} else if (x < 0){
  print("Negative Number")
} else
  print("zero")
```

```{r}
a <- c(5,7,2,9)
ifelse(a %% 2==0, "even", "odd")
```
b. for loops

```{r}
z <- c(2,5,3,9,8,11,6)
count <- 0
for (val in z) {
  if (val %% 2==0) count = count +1
}
print(count)
```
c. while statement

```{r}
i <- 1
while (i < 5) {
  print(i)
  i= i+1
}
```
d. Interruptions

```{r}
x <- 1:5

for (val in x) {
  #print
  if (val == 3){
    #print(val)
    break
  }
  print(va)
}
```

```{r}
x <- 5
for (val in x) {
  if (val ==3) {
    next
  }
  print(val)
}
```
next means "skip"
```{r}
x <- 1
repeat{
  print(x)
  x = x+1
  if (x ==20) {
    break

  }
}
```
## Randomization and dataset management (short)

```{r}
B <- matrix(
  c(2,4,3,1,5,87),
  nrow=3,
  ncol=2
)

B
```
```{r}
t(B)
```

```{r}
C <- matrix (
  c(7,4,2),
)
```
 you missed some here
 
 b. randomizations
```{r}
#norm(n,mean,sd)
```


```{r}
#sample(x, size, replace = FALSE, prob = NULL)
```
```{r}
ndist <- rnorm(100, 50, 10)
ndist

mean(ndist)
sd(ndist)

hist(ndist)
```
```{r}
sample(ndist, 10)
```
  
```{r}
test <- 1:100
test
```
```{r}
sample(test, 50, replace=TRUE)
```

